---
title: "CATEDisplays: Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CATEDisplays-demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width=8, fig.height=5) 
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>"
)
```

## Getting Started: Generating Data

__Elena Badillo-Goicoechea, Ryoko Susukida, and Elizabeth A. Stuart__

2023-11-19

## 1. Introduction

Personalized treatment decisions, which hold great potential for improving health outcomes for patients of varying characteristics, involve the assessment and estimation of individualized or conditional treatment effects. 

However, using data to make these decisions in a reliable, precise, and generalizable way is challenging when sample sizes are low and/or are not representative of the target population. Leveraging the information provided by multiple studies with unconfounded treatment assignment allows for combining datasets to improve the power to estimate heterogeneous treatment effects.

CATEDisplays is designed to quickly estimate the conditional average treatment effect from one or more trials, and provide descriptive and formal diagnostics to assess the presence of treatment effect heterogeneity.

The functions in this package are wrapped in the main function [displayCATE()]() but can run separately. These let the user seamlessly: 

1) Estimate the conditional average treatment effect (CATE) coming from one or more sites and/or trials
2) Combine data from multiple trials (optionally).
3) Test for the presence of treatment effect heterogeneity, using multiple diagnostics, including agnostic test for hetergeneity, distribution of the CATE by subgroups of interest with confidence intervals, and a doubly robust best linear projection of the CATE, provided user-specified covariates.
4) Visualize results

Overall, the procedure This package is mainly designed for researchers evaluating evidence coming from multiple trials and interested in inspecting potentially differential treatments effects across different subpopulations.

To estimate causal forests and the best linear projection, CATEDisplays uses the [grf](https://cran.r-project.org/web/packages/grf/index.html) R package.


This vignette will guide the reader, step-by-step, on CATEDisplays basic functionalities, with simulated datasets. Technical details on each of the package's main functions and on the sample dataset is included in the documentation, which the reader is encouraged to consult, if necessary.


## 2. Package and data set-up

First, if needed, install the CATEDisplays package with the help of devtools:

```{r}
if(!require(devtools)){
    install.packages("devtools", repos = "http://cran.us.r-project.org")
}

if(!require(CATEDisplays )){
    devtools::install_github("StuartRA/CATEDisplays ")
}

```

And load the package by calling its library:

```{r}
library(CATEDisplays)
```

Now let's generate a dataset for 5 trials that satisfies the following basic structure (required for the package to run seamlessly):

1) One dataframe per study/trial, with varying sample sizes (250, 50, 450, 350, 150)
2) If more than one study or trial is included, all the specified covariates of interest (e.g. demographics) must be available for each study/trial and coded in the same manner.
3) Numerical outcome variable.

Although technically not required, if the user wants to merge data coming from multiple trials, it is highly recommended that the treatment and outcome have a similar meaning across studies.

```{r, results='asis'}

df1 <- data.frame(matrix(NA, nrow = 250, ncol = 1))
df2 <- data.frame(matrix(NA, nrow = 50, ncol = 1))
df3 <- data.frame(matrix(NA, nrow = 450, ncol = 1))
df4 <- data.frame(matrix(NA, nrow = 350, ncol = 1))
df5 <- data.frame(matrix(NA, nrow = 150, ncol = 1))
```

Generate a categorical covariate ("Sex"):

```{r, results='asis'}
df1$sex <- as.factor(sample(c('Male', 'Female'), 250, replace=TRUE))
df2$sex <- as.factor(sample(c('Male', 'Female'), 50, replace=TRUE))
df3$sex <- as.factor(sample(c('Male', 'Female'), 450, replace=TRUE))
df4$sex <- as.factor(sample(c('Male', 'Female'), 350, replace=TRUE))
df5$sex <- as.factor(sample(c('Male', 'Female'), 150, replace=TRUE))
```

Generate a continuous covariate ("Age"):

```{r, results='asis'}
df1$Age <- unlist(lapply(rnorm(250, 40, 15), function (x) {max(sample(c(18:20)), round(x, 2))}))
df2$Age <- unlist(lapply(rnorm(50, 30, 15), function (x) {max(sample(c(18:20)), round(x, 2))}))
df3$Age <- unlist(lapply(rnorm(450, 25, 15), function (x) {max(sample(c(18:20)), round(x, 2))}))
df4$Age <- unlist(lapply(rnorm(350, 20, 15), function (x) {max(sample(c(18:20)), round(x, 2))}))
df5$Age <- unlist(lapply(rnorm(150, 40, 15), function (x) {max(sample(c(18:20)), round(x, 2))}))
```
Generate a categorical covariate ("Educational level"):
```{r, results='asis'}
df1$educ <- as.factor(sample(c('Less than HS', 'HS', 'College', 'Graduate'), 250, replace=TRUE))
df2$educ <- as.factor(sample(c('Less than HS', 'HS', 'College', 'Graduate'), 50, replace=TRUE))
df3$educ <- as.factor(sample(c('Less than HS', 'HS', 'College', 'Graduate'), 450, replace=TRUE))
df4$educ <- as.factor(sample(c('Less than HS', 'HS', 'College', 'Graduate'), 350, replace=TRUE))
df5$educ <- as.factor(sample(c('Less than HS', 'HS', 'College', 'Graduate'), 150, replace=TRUE))
```

Generate a categorical covariate ("Race"):
```{r, results='asis'}
df1$race <- as.factor(sample(c('White', 'Black', 'Asian', 'Hispanic', 'Other'), 250, replace=TRUE))
df2$race <- as.factor(sample(c('White', 'Black', 'Asian', 'Hispanic', 'Other'), 50, replace=TRUE))
df3$race <- as.factor(sample(c('White', 'Black', 'Asian', 'Hispanic', 'Other'), 450, replace=TRUE))
df4$race <- as.factor(sample(c('White', 'Black', 'Asian', 'Hispanic', 'Other'), 350, replace=TRUE))
df5$race <- as.factor(sample(c('White', 'Black', 'Asian', 'Hispanic', 'Other'), 150, replace=TRUE))
```

Generate a continuous covariate ("Alcohol Addiction Severity Index"):

```{r, results='asis'}
df1$ASIalc <- rnorm(250, 0, 0.55)
df2$ASIalc <- rnorm(50, 0, 0.25)
df3$ASIalc <- rnorm(450, 0, 0.15)
df4$ASIalc <- rnorm(350, 0, 0.25)
df5$ASIalc <- rnorm(150, 0, 0.35)
```

Generate a continuous outcome variable:

```{r, results='asis'}
df1$outc <- rnorm(250, 0.6, 0.75)
df2$outc <- rnorm(50, 0.6, 0.75)
df3$outc <- rnorm(450, 0.6, 0.75)
df4$outc <- rnorm(350, 0.6, 0.75)
df5$outc <- rnorm(150, 0.6, 0.75)
```

Generate a binary treatment assignment variable:

```{r, results='asis'}
df1$tx <- sample(c(1, 0), 250, replace=TRUE, prob = c(0.5, 0.5))
df2$tx <- sample(c(1, 0), 50, replace=TRUE, prob = c(0.25, 0.75))
df3$tx <- sample(c(1, 0), 450, replace=TRUE, prob =c(0.3, 0.7))
df4$tx <- sample(c(1, 0), 350, replace=TRUE, prob = c(0.5, 0.5))
df5$tx <- sample(c(1, 0), 150, replace=TRUE, prob = c(0.45, 0.55))
```

Finally, put the single trial dataframes into a list:

```{r, results='asis'}
dfList <- list(df1, df2, df3, df4, df5)
```

## 3. Run main wrapper function:

### Option 1: Not combining trials

Here [cateDisplay()]() will estimate a casual forest and assess treatment effect heterogeneity separately for each single trial.
 
```{r, results='asis'}
catesDisplay <- CATEDisplays::displayCATE(dfList,
                           outCol="outc",
                           txCol="tx",
                           covList= c("age", "race", "educ", "sex", "ASIalc"),
                           blpredList=NULL,
                           combine=F,
                           verbose=T,
                           ci=0.95, nTrees=10000, seedN=7203)
```

Inspect output elements:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
cateOutput <- catesDisplay$cateOutput

# ATE:
cateOutput$ATE

# HTE:
cateOutput$testHTE

# Augmented dataset:
cateOutput$cateDF

# Best linear projection:
jtools:summ(cateOutput$BLP)
```

Save figures with visualizations:

```{r, results='asis', echo=FALSE, message=FALSE, warning=FALSE}

pdf("alltausByStudy.pdf",  width=10, height=6)
  print(catesDisplay$vizTauHat)
dev.off()

pdf("tausBySubgroupAndStudy.pdf",  width=10, height=6)
  print(catesDisplay$vizSubgroupCATE)
dev.off()

pdf("blpByStudy.pdf", width=10, height=6)
  print(catesDisplay$vizBLP)
dev.off()
```


### Option 2: Combining trials

Here [cateDisplay()]() will first combine the data from all trials, and then estimate the CATE casual forest (including a study indicator) and assess treatment effect heterogeneity. As noted above, although technically not required, when merging data coming from multiple trials, it is highly recommended that the treatment and outcome have a similar meaning across studies.

```{r, results='asis',echo=FALSE, message=FALSE, warning=FALSE}
catesDisplayCombined <- CATEDisplays::displayCATE(dfList,
                                  outCol="outc",
                                  txCol="tx",
                                  covList= c("age", "race", "educ",
                                             "sex", "ASIalc"),
                                  blpredList=NULL,
                                  combine=T,
                                  ci=0.95,
                                  verbose=T,
                                  nTrees=10000,
                                  seedN=7203)
```

Again, we can inspect the elements of the main output, and save resulting visuals.

```{r, results='asis',echo=FALSE, message=FALSE, warning=FALSE}

pdf("alltausCombined.pdf",  width=10, height=6)
  print(catesDisplayCombined$vizTauHat)
dev.off()

pdf("tausBySubgroupCombined.pdf",  width=10, height=6)
  print(catesDisplayCombined$vizSubgroupCATE)
dev.off()


pdf("blpCombined.pdf", width=10, height=6)
  print(catesDisplayCombined$vizBLP)
dev.off()

```
